Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-11-19 response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1\vskip 5mm}
\newcommand{\refb}[1]{\textbf{R2:} #1\vskip 5mm}
\newcommand{\editor}[1]{\textbf{Editor:} #1\vskip 5mm}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}}

@@

\refa{The manuscript by Barrat-Charlaix et al. discusses the problem of
predictability of mutations in seasonal influenza viruses. It presents
data analysis of frequency trajectories in H1N1 and H3N2 lineages and
simulations of different selection scenarios.
It is an interesting revisit to predictions in influenza, however I
have some major concerns about the analysis and formulation of the
conclusions that should be addressed:}

\refa{1. The prediction problem is defined for each mutation: based on the
tracked frequency trajectory, can the future (fixation, loss, or
polymorphism) of the mutation be predicted. Such formulation has been
previously proposed by Illingworth and Mustonen, (eg. Genetics 2011,
Plos Pathogens 2012). By averaging over all amino-acid substitutions,
the authors show that the fates of mutations are not determined by the
value of the starting frequency.}

\refa{1.1. Mutations in influenza are highly nested, with a substantial
hitch-hiking, and no attempt is made to disentangle such dependencies when counting the mutations.}

\response{This is a good suggestion, and we agree with reviewer 1 that our original manuscript lacked a method to disentangle nested mutations. \\
We added a new section in the Supplementary Material where we attempt to cluster together trajectories of mutations that partly appear on the same strains. 
Trajectories corresponding to mutations always or often appearing on the same strains are then counted as one "effective" trajectory. 
However, this new way of counting mutations does not significantly change our results, and as a consequence we left the figures of main text unchanged.}

@@

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\responsedraft{
	\begin{itemize}
		\item Illingworth\&Mustonen do not do prediction. They fit complete frequency trajectories using two models (no linkage and linkage), and find that only one of the two is able to deliver a good fit. The conclusion is that linkage is a dominant effect for shape of trajectories. 
		\item We could indeed make an attempt at modeling the initial shape of a frequency trajectory until the point where a prediction (fixation/loss) is made. However, we feel that a strong point of our approach is to be modeling-free. 
		\item Simulations of populations evolving without recombination show that even without modeling, the fixation probability of a rising trajectory is higher than its frequency at the time when the prediction is made. Our claim is only that observations in A/H3N2 influenza differ both from what is typically expected the theory of evolving populations and from simulations. In particular, we do claim not that the trajectories are intrisically unpredictable. 
	\end{itemize}
}

@@

\refb{1. The authors focus on one feature at a time (frequency, epitope status, LBI …). It is interesting to see that each of these is not very predictive of future frequency / prob of fixation. However, I think the obvious next step is to see whether a combination of many features could do a better job of predicting. I am not sure why the authors don’t try to fit a model that takes into account all information they have about sites (say, type of AA change, location in the gene, current frequency etc) and see if a ML model is able to make predictions. }

\refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
\responsedraft{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}. 
The answer is that yes it's expected that vaccine influence frequency trajectories to some extent, but it is very hard to measure in practice. 
The above paper attempted to measure this, and did not find clear-cut results.}

### 2020-11-19 article.tex
Add markers -- Change Fig 1.

The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.  
In the short term, we found that on average the direction of trajectory does not persist for longer than a few months. 
Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''. 
This suggests that selective sweeps are not representative of typical trajectories.

>>>>

The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.  
In the short term, we found that on average the direction of trajectory does not persist for longer than a few months. 
Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''. 
This suggests that selective sweeps are not representative of typical trajectories. \pierre{We could cite the Illingworth paper here. They do show that as well, in a different way. The fact many trajectories are not sweeps is expected when there is strong linkage. What cannot be easily explained by linkage only is the complete lack of inertia on average (at least I think?), and the $P_{fix}(f)=f$. }
